---
title: "Name-blind variable-role detection with rolescry"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Name-blind variable-role detection with rolescry}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(rolescry)
set.seed(1)
```

## The problem: names lie, data does not

Real datasets arrive with column names that are missing, misleading, in the
wrong language, or simply wrong: a column called `category_code` holding
continuous lab values, a `gender` column that is actually a free numeric
measurement, an outcome buried under an opaque `v7`. Any tool that decides a
column's statistical *role* from its **name** inherits every one of those lies.

`rolescry` decides roles from the **data signature** instead -- the guiding
principle is *Data inspice, non nomen* ("inspect the data, not the name").
Renaming every column to `col_1, col_2, ...` does not change a single role
assignment. This is the **turnusol** (litmus) invariant, and it is the
package's keystone test.

## A worked example

```{r example}
set.seed(1)
pre  <- rnorm(100, 10, 2)          # a measurement, before ...
d <- data.frame(
  arm  = rep(c("control", "treatment"), each = 50),  # a balanced 2-level grouping
  pre  = pre,
  post = pre + rnorm(100, 1, 1)                       # ... and again after (paired with pre)
)

res <- detect_roles(d)
res
summary(res)
```

The same call on the name-stripped twin yields the same roles by position:

```{r blind}
d_blind <- setNames(d, paste0("col_", seq_along(d)))
pos <- function(r, dat) match(r$roles$paired_pairs$columns, names(dat))
identical(pos(detect_roles(d), d), pos(detect_roles(d_blind), d_blind))
```

## How a role is scored

Each column is first **typed by value** (`continuous`, `binary`, `categorical`,
`ID`), never by name. Candidate roles are then scored by signatures that capture
the statistical shape a role implies -- correlation and distributional overlap
for paired measurements, Bland-Altman bias and intraclass correlation for
agreement, event-rate and right-skew for survival, inter-item correlation and a
Cronbach-alpha proxy for scale items, and so on. Every score is a transparent
sum of named components you can inspect:

```{r breakdown}
res$roles$paired_pairs$components[[1]]
```

### Shannon entropy

For a categorical column with level proportions \(p_1, \dots, p_k\), the
normalized Shannon entropy

\[ H_{\text{norm}} = \frac{-\sum_i p_i \log_2 p_i}{\log_2 k} \in [0, 1] \]

measures how balanced the levels are. A grouping variable (treatment vs control)
has high entropy (near-balanced); a near-constant flag has entropy near zero.
Entropy drives both the value classifier and the group-balance signal.

### Normalized mutual information

To ask -- name-blind -- whether a candidate grouping actually *carries
information about* an outcome, `rolescry` uses normalized mutual information:

\[ \text{NMI}(X, Y) = \frac{I(X; Y)}{\min\{H(X),\, H(Y)\}} \in [0, 1], \]

which is 0 for independent variables and 1 for a deterministic association, and
is comparable across variables with different numbers of levels. It is exposed
directly:

```{r nmi}
set.seed(2)
g <- sample(c("A", "B", "C"), 300, replace = TRUE)
y <- ifelse(g == "A", "event", sample(c("event", "none"), 300, replace = TRUE))
compute_nmi(g, y)          # > 0: g informs y
compute_nmi(g, sample(g))  # ~ 0: shuffled -> independent
```

## The optional, capped name bonus

Names are not *useless* -- they are just untrustworthy. When you do trust them,
pass a keyword dictionary via `name_bonus`. Names then act only as a small,
**capped** tie-breaker (at most a +10 point nudge, i.e. <= 10% of the selection
score); the mathematical signature still dominates (>= 90%), the relationship
enforced by `score_gap_ok()`.

Here two columns both look like groupings; the data alone picks the perfectly
balanced `site`, but a keyword dictionary nudges the choice to the intended
treatment arm:

```{r namebonus}
set.seed(4)
clin <- data.frame(
  site    = rep(c("north", "south"), length.out = 160),  # perfectly balanced (the math pick)
  treated = sample(c("no", "yes"), 160, replace = TRUE, prob = c(0.62, 0.38))
)
detect_roles(clin)$roles$group_var$columns                                  # data alone -> "site"
detect_roles(clin, name_bonus = rolescry_default_name_bonus())$roles$group_var$columns  # name nudge -> "treated"
```

## Header-aware loading

`read_data()` reads a file with the header row found by the same
information-theoretic scorer (`detect_header()`), so messy exports with title
rows or merged cells still load with sensible column names. Delimited text works
with base R; spreadsheet and statistical formats use optional packages and
degrade gracefully if they are not installed.

## Attribution

`rolescry` is derived from Boynukara, C. (2026). *MDStatR* (v2.1.0 Veritas).
Zenodo. <https://doi.org/10.5281/zenodo.20707791>. Run `citation("rolescry")`
to cite the package and its parent engine.
